Particle MCMC for Multivariate Stochastic Differential Equations with **`msde`**

### 2018-03-22 inst/doc/msde-pmcmc.Rmd
recompiled msde-pmcmc.Rmd

---
title: "Particle MCMC for Multivariate Stochastic Differential Equations with **`msde`**"
author: "Martin Lysy, Feiyu Zhu"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
bibliography: references.bib
csl: taylor-and-francis-harvard-x.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Particle MCMC with msde}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
\newcommand{\Ep}{\mathrm{E}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\Cov}{\mathrm{Cov}}
\newcommand{\idm}{\mathrm{I}}
\newcommand{\bm}[1]{\boldsymbol{#1}}

@@

Currently, `sde.pmcmc` in **`msde`** only implements the *particle marginal Metropolis-Hastings* (PMMH) sampler in the particle MCMC toolbox which will be used for our later simulation study. PMMH is an approximation of an ideal *pseudo marginal Metropolis-Hastings* sampler (see @pseudoMCMC) for sampling from $p(\theta, x_{1:T} | y_{1:T})$. It utilizes the following proposal density 
$$
q((\theta^*, x^*_{1:T}) | (\theta, x_{1:T})) := q(\theta^* | \theta) p_{\theta^*}(x^*_{1:T} | y_{1:T})
$$
where the proposed $x^*_{1:T}$ is perfectly *adapted* to the proposed $\theta^*$. The proposal density $q(\theta^* | \theta)$ should be specified by users. In `sde.pmcmc`, a very basic IID-Metropolis proposal is implemented. Samples from $p_{\theta^*}(x^*_{1:T} | y_{1:T})$ can be obtained from an SMC algorithm targeting $p_{\theta^*}(x^*_{1:T} | y_{1:T})$. In fact we do not need the analytic expression of $p_{\theta^*}(x^*_{1:T} | y_{1:T})$ to calculate the acceptance ratio since the ratio can be transformed as
$$
\frac{p(\theta^*, x^*_{1:T} | y_{1:T})}{p(\theta, x_{1:T} | y_{1:T})} \frac{q((\theta, x_{1:T}) | (\theta^*, x^*_{1:T}))}{q((\theta^*, x^*_{1:T}) | (\theta, x_{1:T}))} = \frac{p_{\theta^*}(y_{1:T}) p(\theta^*)}{p_\theta(y_{1:T})p(\theta)} \frac{q(\theta | \theta^*)}{q(\theta^* | \theta)}
$$
In the above formula, $p_\theta(y_{1:T})$ can be replaced by its estimated expression $\hat{p}_\theta(y_{1:T})$
$$
\hat{p}_\theta(y_{1:T}) := \hat{p}_\theta(y_1) \prod_{n=2}^T \hat{p}_\theta(y_n | y_{1:n-1})
$$
which are approximated by the unnormalized particle weights (not the incremental weights) at time $T$, i.e.
$$
\hat{p}_\theta(y_{1:T}) = \frac{1}{N} \sum_{i=1}^N w(X_{1:T}^i)
$$
where $N$ is the pre-specified total number of particles.

@@

### Posterior sampling via particle MCMC
Let's set the number of particles as `100` and use just IID-Metropolis proposal for updating the unknown parameter $\Lambda_1$. Here our aim is to show that even "off-the-shelf" choice can give satisfactory result.
```{r pmcmc, cache = TRUE}
npart <- 100
mwg.sd <- 0.3 # for convenience, just use a fixed jump size instead of an adaptive one
ppost <- sde.pmcmc(model = bmod, binit, theta0, fixed.params, nsamples, npart, dT,
                   resample = "multi", threshold = 0.5, mwg.sd)
```
Then we can check the acceptance rate.
```{r accept}
# check the acceptance rate
accept <- ppost$accept
print(accept)
```
The acceptance rate is 94.6% which is close to the optimal 100%.

>>>>

---
title: "Particle MCMC for Multivariate Stochastic Differential Equations with **`msde`**"
author: "Martin Lysy, Feiyu Zhu"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
bibliography: references.bib
csl: taylor-and-francis-harvard-x.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Particle MCMC with msde}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
\newcommand{\Ep}{\mathrm{E}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\Cov}{\mathrm{Cov}}
\newcommand{\idm}{\mathrm{I}}
\newcommand{\bm}[1]{\boldsymbol{#1}}
\newcommand{\N}{\mathcal{N}}

@@

Currently, `sde.pmcmc`^[Since `sde.pmcmc` is still under development, it is currently not available for users of **`msde`**.] in **`msde`** only implements the *particle marginal Metropolis-Hastings* (PMMH) sampler in the particle MCMC toolbox which will be used for our later simulation study. PMMH is an approximation of an ideal *pseudo marginal Metropolis-Hastings* sampler (see @pseudoMCMC) for sampling from $p(\theta, x_{1:T} | y_{1:T})$. It utilizes the following proposal density
$$
q((\theta^*, x^*_{1:T}) | (\theta, x_{1:T})) := q(\theta^* | \theta) p_{\theta^*}(x^*_{1:T} | y_{1:T})
$$
where the proposed $x^*_{1:T}$ is perfectly *adapted* to the proposed $\theta^*$. The proposal density $q(\theta^* | \theta)$ should be specified by users. In `sde.pmcmc`, a very basic IID-Metropolis proposal is implemented. Samples from $p_{\theta^*}(x^*_{1:T} | y_{1:T})$ can be obtained from an SMC algorithm targeting $p_{\theta^*}(x^*_{1:T} | y_{1:T})$. In fact we do not need the analytic expression of $p_{\theta^*}(x^*_{1:T} | y_{1:T})$ to calculate the acceptance ratio since the ratio can be transformed as
$$
\frac{p(\theta^*, x^*_{1:T} | y_{1:T})}{p(\theta, x_{1:T} | y_{1:T})} \frac{q((\theta, x_{1:T}) | (\theta^*, x^*_{1:T}))}{q((\theta^*, x^*_{1:T}) | (\theta, x_{1:T}))} = \frac{p_{\theta^*}(y_{1:T}) p(\theta^*)}{p_\theta(y_{1:T})p(\theta)} \frac{q(\theta | \theta^*)}{q(\theta^* | \theta)}
$$
In the above formula, $p_\theta(y_{1:T})$ can be replaced by its estimated expression $\hat{p}_\theta(y_{1:T})$
$$
\hat{p}_\theta(y_{1:T}) := \hat{p}_\theta(y_1) \prod_{n=2}^T \hat{p}_\theta(y_n | y_{1:n-1})
$$
which are approximated by the unnormalized particle weights (not the incremental weights) at time $T$, i.e.
$$
\hat{p}_\theta(y_{1:T}) = \frac{1}{N} \sum_{i=1}^N w(X_{1:T}^i)
$$
where $N$ is the pre-specified total number of particles.

@@

### Posterior sampling via particle MCMC
Let's set the number of particles as `100` and just use an adaptive random walk proposal for updating the unknown parameter $\Lambda_1$. Here our aim is to show that even "off-the-shelf" choice can give satisfactory result.
```{r source, echo = FALSE}
source("sde.pmcmc.R")
```
```{r pmcmc, cache = TRUE}
npart <- 100
mwg.sd <- 1
delta <- .618
ppost <- sde.pmcmc(model = bmod, binit, theta0, fixed.params, hyper,
                   nsamples, npart, dT,
                   resample = "multi", threshold = 0.5, mwg.sd, delta)
```
In general we have the following proposal:
$$
\theta_{j, prop} \sim \N(\theta_{j, curr}, \sigma^2_{n,j})
$$
where $\theta_{j,prop}$ is the proposed (new) $j$-th parameter and $\theta_{j,curr}$ is the current (or old) $j$-th parameter. Besides, the adaptive standard deviation $\sigma_{n+1,j}$ at the $n+1$-th iteration for the $j$-th parameter is determined by
$$
\sigma_{n+1,j} = \exp\left(\log(\sigma_{n,j}) \pm \frac{\delta}{n} \right).
$$
Based on the above adaptive rule, we increase/decrease $\sigma_{n,j}$ at each step if the previous draw was accepted/rejected. The amount of adaptation will goes to $0$ as the number $n$ of iterations goes large. This kind of adaptation can preserve the MCMC stationary distribution.

Then we can check the acceptance rate.
```{r accept}
# check the acceptance rate
accept <- ppost$accept
print(accept)
```
The acceptance rate is 50.29% which is close to the optimal 45%.
